CMP_monocle

### 2021-12-13 monocle/IndividualPops_monocle.Rmd
Completed initial trajectory inference for CMP via IndividualPops script

---
title: "CMP_monocle"
output: html_notebook
---
# Setup
## Define notebook variables
```{r}
working.genome <- "mm10"
umi.cutoff <- 200
data.path <- "/Users/heustonef/Desktop/10XGenomicsData/msAggr_scRNASeq/IndividualPops/"
cell.pop <- "CMP"
```

## Load libraries
```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Desktop/10XGenomicsData/msAggr_scRNASeq/Monocle/")
library(monocle3)
library(ggplot2)
library(dplyr)
```
## Source functions
```{r}
source("~/Desktop/10XGenomicsData/msAggr_scRNASeq/RFunctions/ColorPalette.R")
```
# Load datasets
## Provide list of datasets
After messing with this a bit, I've decided the easiest thing to do is integrate the data from the seurat object. This object already has the metadata and gene_short_names that I will need, and based on monocle3 documentation may be the easiest way to make sure everything gets where it needs to be.
```{r, eval=FALSE}
# Don't run
library(Seurat)
data.path <- 
seurat.object <- readRDS("../msAggr_seurat/msAggr_seurat_dim26.RDS")

cell_metadata <- seurat.object@meta.data

```

>>>>

---
title: "CMP_monocle"
output: html_notebook
---
# Setup
## Define notebook variables
```{r}
working.genome <- "mm10"
umi.cutoff <- 200
data.path <- "/Users/heustonef/Desktop/10XGenomicsData/msAggr_scRNASeq/IndividualPops/"
cell.pop <- "CMP"
seurat.var.pct <- 90
n.neighbors <- 20

# Setting a seed to ensure reproducible data
seed.value <- 10
```
## Notebook variable definitions
```{r}
seurat.rds <- paste0(cell.pop, "Subpop_pctVar", seurat.var.pct, ".RDS")
projectName <- paste0(cell.pop, "Subpop_monocle" )
set.seed(seed.value)
```

## Load libraries
```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_knit$set(root.dir = "~/Desktop/10XGenomicsData/msAggr_scRNASeq/Monocle/")
library(monocle3)
library(ggplot2)
library(dplyr)
```
## Source functions
```{r}
source("~/Desktop/10XGenomicsData/msAggr_scRNASeq/RFunctions/ColorPalette.R")
```
## Store session info
```{r}
sessionInfo.filename <- paste0(projectName, "_sessionInfo.txt")
sink(sessionInfo.filename)
sessionInfo()
sink()
```

# Load datasets
## Provide list of datasets
After messing with this a bit, I've decided the easiest thing to do is integrate the data from the seurat object. This object already has the metadata and gene_short_names that I will need, and based on monocle3 documentation may be the easiest way to make sure everything gets where it needs to be.
```{r}
library(Seurat)
seurat.object <- readRDS(file.path(data.path, seurat.rds))
seurat.xprsn <- GetAssayData(seurat.object, slot = "counts", assay = "RNA")
cell_metadata <- seurat.object@meta.data[colnames(seurat.xprsn),]
gene_metadata <- data.frame(gene_short_name = rownames(seurat.xprsn))
rownames(gene_metadata) <- rownames(seurat.xprsn)
cds <- new_cell_data_set(expression_data = seurat.xprsn, 
												 cell_metadata = cell_metadata[colnames(seurat.xprsn), ], 
												 gene_metadata = gene_metadata)
rm(seurat.object, seurat.xprsn, cell_metadata, gene_metadata)
```
Look at object
```{r}
cds
```
## Save CDS object
```{r}
saveRDS(cds, file = paste0(projectName, ".RDS"))
```
# Data analysis
## Load saved object
```{r}
cds <- readRDS(paste0(projectName, ".RDS"))
```

## Preprocessing
```{r}
cds <- preprocess_cds(cds = cds, metcdshod = "PCA", num_dim = 50, verbose = TRUE)
```
```{r}
plot_pc_variance_explained(cds) + ggtitle("Proportion of variance explained (note not stdev)")
```
## calculate proportion of variance explained
```{r}
pct.list <- c(80, 85, 90, 95)
for (pct in pct.list){
	print(paste("Num pcs for", pct, "variance:", length(which(cumsum(cds@preprocess_aux$prop_var_expl*100) <= pct))))
}
```

## Clustering analysis
### Cluster with UMAP.
UMAP is based on uwot's `umap` function
```{r}
cds <- reduce_dimension(cds = cds, max_components = 2, reduction_method = "UMAP", preprocess_method = "PCA", umap.n_neighbors = n.neighbors)
```
plot umap. Can color cells based on any `colData(cds)` column (essentially metadata columns)
```{r}
plot_cells(cds = cds, reduction_method = "UMAP", color_cells_by = "RNA_snn_res.1", show_trajectory_graph = FALSE, cell_size = .7, group_label_size = 5) + scale_color_manual(values = color.palette)
```
### Cluster wiht tSNE
Note that this is stored in a different slot than UMAP and will not overwrite those data
```{r}
cds <- reduce_dimension(cds = cds, max_components = 2, reduction_method = "tSNE", preprocess_method = "PCA")
```
Plot tSNE. Must specifiy slot to plot
```{r}
plot_cells(cds = cds, reduction_method = "tSNE", color_cells_by = "RNA_snn_res.1", show_trajectory_graph = FALSE, cell_size = .7, group_label_size = 5) + scale_color_manual(values = color.palette)
```
## Save CDS object
```{r}
saveRDS(cds, file = paste0(projectName, ".RDS"))
```
### Check for batch effects
Function `align_cds` tries to remove batch effects using mutual nearest neighbor alignment (package `batchelor`, Aaron Lun). Will test this on a different object.  
Can specify alignment variable using parameter `alignment_group`. Note that only 1 variable can be specified
```{r}
# cds.aligned <- align_cds(cds = cds, preprocess_method = "PCA", alignment_group = "nCount_RNA")
```
*Failboat. Requires too much memory on laptop.*
### Perform clustering analysis
Monocle uses "community detection" (Leiden method and Louvain method).  
Note that setting `resolution = NULL` and `cluster_method = "leiden"`, then resoltuion is determined automatically (actually I think it's just set to 0.0001, based on code)...
```{r}
# read RDS
cds <- readRDS(file = paset0(projectName, ".RDS"))
```

```{r}
res.list <- c(1e-4, 5e-4, 1e-3, 2e-3)
# reduction.method <- "UMAP"
# cluster.method <- "leiden"

for(res in res.list){
	# Cluster cells at given resolution
	cds <- cluster_cells(cds = cds, 
											 reduction_method = "UMAP", 
											 k = n.neighbors, 
											 cluster_method = "leiden", 
											 resolution = res, 
											 num_iter = 10, 
											 random_seed = seed.value,
											 verbose = TRUE)
	# Copy cluster information to metadata in colData
	coldata.res <- paste0("umap_res_", res)
	colData(cds)[coldata.res] <- clusters(cds)
}
for(colID in colnames(colData(cds))){
	if(grepl("umap_res", colID)){
		cat("cluster count for", colID, ":", length(levels(colData(cds)[,colID])), "\n")
	}
}

```
## save umap object
```{r}
saveRDS(cds, file = paste0(projectName, "-UMAP.RDS"))
```
## read object
```{r}
cds <- readRDS(paste0(projectName, "-UMAP.RDS"))
```

```{r}
for(colID in colnames(colData(cds))){
	if(grepl("umap_res", colID)){
		if(length(levels(colData(cds)[,colID])) <= length(color.palette)){
			cluster_plot <- plot_cells(cds = cds, reduction_method = "UMAP", 
																 color_cells_by = colID, show_trajectory_graph = FALSE, 
																 cell_size = 1, group_label_size = 5, label_groups_by_cluster = FALSE) + 
				scale_color_manual(values = color.palette) +
				ggtitle(paste0(projectName, "-", colID))
			plot(cluster_plot)
		} else {
			cluster_plot <- plot_cells(cds = cds, reduction_method = "UMAP", 
																 color_cells_by = colID,  show_trajectory_graph = FALSE, 
																 cell_size = 1, group_label_size = 5, label_groups_by_cluster = FALSE) + 
				ggtitle(paste0(projectName,  "-", colID))
			plot(cluster_plot)

		}
		# save plot
		png(filename = paste0(projectName,  "-", colID, ".png"), height= 800, width = 800)
		plot(cluster_plot)
		dev.off()
	}
}
```

### Explore resolution
```{r eval = FALSE}
# library(igraph)
# library(schex)
# library(scater)
# library(scran)
# library(ggrepel)
# # plot_hexbin_meta_shiny(cds, col = "RNA_snn_res.1", action = "majority", min_nbins = 20, max_nbins = 50, dimension_reduction = "UMAP", na.rm = TRUE) 
```
For CMP, let's start exploring with res = 0.001
# Find Marker Genes
randomly guessing to set reference_cells value at 1000 (could also provide a list of cells if, e.g., wanted to specifiy an originating population)
```{r}
marker_test_res <- top_markers(cds, group_cells_by = "umap_res_0.001", reference_cells = 1000, cores = 8)
```
```{r}
marker_test_res
```
KEY:
a data.frame where the rows are genes and the columns are  
* gene_id vector of gene names
* gene_short_name vector of gene short names
* cell_group character vector of the cell group to which the cell belongs
* marker_score numeric vector of marker scores as the fraction expressing scaled by the specificity. The value ranges from 0 to 1.
* mean_expression numeric vector of mean normalized expression of the gene in the cell group
* fraction_expressing numeric vector of fraction of cells expressing the gene within the cell group
* specificity numeric vector of a measure of how specific the gene's expression is to the cell group based on the Jensen-Shannon divergence. The value ranges from 0 to 1.
* pseudo_R2 numeric vector of pseudo R-squared values, a measure of how well the gene expression model fits the categorical data relative to the null model. The value ranges from 0 to 1.
* marker_test_p_value numeric vector of likelihood ratio p-values
* marker_test_q_value numeric vector of likelihood ratio q-values  
```{r}
top_pseudoR2_markers <- marker_test_res %>%
	filter(fraction_expressing >= 0.10) %>%
	group_by(cell_group) %>%
	top_n(3, pseudo_R2)
top_pseudoR2_markers <- unique(top_pseudoR2_markers %>% pull(gene_id))
my_plot <- plot_genes_by_group(cds, top_pseudoR2_markers, group_cells_by = "umap_res_0.001", ordering_type = "maximal_on_diag", max.size = 10)
plot(my_plot)
png(filename = paste0(projectName, "-map_res_0.001", "-pseudoR2markers.png"), height = 1200, width = 800)
plot(my_plot)
dev.off()
```

# Auto Annotate with Garnett
## Start with marker_test_res
```{r}
# Require that markers have at least JS specificty score > 0.5 and
# be significant in the logistic test for identifying their cell type:
garnett_markers <- marker_test_res %>%
                        filter(marker_test_q_value < 0.01 & specificity >= 0.5) %>%
                        group_by(cell_group) %>%
                        top_n(5, marker_score)
# Exclude genes that are good markers for more than one cell type:
garnett_markers <- garnett_markers %>% 
                        group_by(gene_short_name) %>%
                        filter(n() == 1)
# Generate garnett marker file
generate_garnett_marker_file(garnett_markers, file = paste0(projectName, "-garnettMarkers.txt"))
```

__Note that this file can be manually modified to add/remove classifier markers
__Abandoning this!__ 1) I'm getting distracted and 2) Will have to play with resolution to get marker genes for more than 3 clusters

# Trajectory Inference
First make sure correct resolution is represented in cluster slot
```{r}
cds <- cluster_cells(cds = cds, 
										 reduction_method = "UMAP", 
										 k = n.neighbors, 
										 cluster_method = "leiden", 
										 resolution = 0.001, 
										 num_iter = 10, 
										 random_seed = seed.value,
										 verbose = TRUE)
```
Fit principal graph. Set `use_partition = TRUE` to not require all cells be connected on a single graph
```{r}
cds <- learn_graph(cds, use_partition = TRUE, verbose = TRUE)
```
## save result
```{r}
saveRDS(cds, file = paste0(projectName, "-umap_res0.001_psdtime.RDS"))
```

```{r}
my_plot <- plot_cells(cds, reduction_method = "UMAP", 
											color_cells_by = "cluster", 
											label_groups_by_cluster = FALSE, label_leaves = FALSE, label_branch_points = TRUE,  
											cell_size = 1, group_label_size = 8, graph_label_size = 5) + 
					scale_color_manual(values = color.palette) +
					ggtitle(paste0(projectName, "-umap_res0.001_psdtime"))
plot(my_plot)

png(filename = paste0(projectName, "-umap_res0.001_psdtime.png"), height = 800, width = 800)
plot(my_plot)
dev.off()
```

Will need to complete msAggr to order cells. For example, will pick those CMP cells that cluster with LSK and "earliest" CMPs.
